---
title: "GIS 335 Final Presentation - GeoDa in Collaboration with R"
author: "James Jahraus"
date: "2021-12-13"
output:
 github_document:
  toc: TRUE
---

<!-- Environment Setup -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),
                      tidy=TRUE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


## Presentation Objective

Share with the class the experience of using GeoDa in collaboration with R.

**Key take aways:**

* Demonstrate how GeoDa works in collaboration with R.

* Gain curiosity about using GeoDa for exploratory spatial data analysis.

The exploration is guided by Kolak (2018) Array of Things (aot) Workshop, and L4 from
PennState GEOG 586.


## Scenario to Guide the Exploration

A local Chicago company is developing a temperature web service. The goal is to provide
a temperature map of Chicago every hour with temperature data at every location across the city.
The website will use this data to send users the temperature given their lat lon coordinates.

To create the temperature surface the company has a network of sensors. The aot data set
represents a network of sensors throughout Chicago that collect temperature, humidity, and pressure.

* Sensor **tsys01** collected only temp data in C [TSYS01 - High Accuracy Temperature Sensor](https://www.te.com/commerce/DocumentDelivery/DDEController?Action=showdoc&DocId=Data+Sheet%7FTSYS01%7FA%7Fpdf%7FEnglish%7FENG_DS_TSYS01_A.pdf%7FG-NICO-018)

* Sensor **htu21d** collected temp in C and humidity in RH data [HTU21D-F Temperature + Humidity Sensor](https://www.adafruit.com/product/1899)

* Sensor **bmp180** collected temp in C and pressure in hPa [BMP180 Barometric Pressure/Temperature/Altitude Sensor](https://www.adafruit.com/product/1603)

This data is point data, so making the temperature surface requires interpolation
of the point data to the temperature surface. The deliverable for the web service
is an accurate temperature surface delivered every hour.


## Generate Shapefile for All Sensors

**Choose the Hour for Node Data**


```{r node_data}
# Load Packages
library(tidyverse)
library(lubridate)
library(sp)
library(rgdal)

# Clear Workspace
rm(list = ls())

# Read Data
sensor_data <- read.csv("Data/data.csv.gz")
nodes <- read.csv("Data/nodes.csv")

# Fix timestamp so GeoDa doesn't crash on import
sensor_data$timestamp <- ymd_hms(sensor_data$timestamp)
nodes$start_timestamp <- ymd_hms(nodes$start_timestamp)
nodes$end_timestamp <- ymd_hms(nodes$end_timestamp)

# Filter Sensors by Sensor for hour
fhour <- 14 # filter hour

# tsys01 temperature C
temp_tsys01 <- sensor_data %>%
  filter(sensor == "tsys01") %>%
  filter(parameter == "temperature") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(t_tsys01 = mean(value_hrf))

# htu21d temperature C
temp_htu21d <- sensor_data %>%
  filter(sensor == "htu21d") %>%
  filter(parameter == "temperature") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(t_htu21d = mean(value_hrf))

# htu21d humidity RH
humid_htu21d <- sensor_data %>%
  filter(sensor == "htu21d") %>%
  filter(parameter == "humidity") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(h_htu21d = mean(value_hrf))

# bmp180 temperature C
temp_bmp180 <- sensor_data %>%
  filter(sensor == "bmp180") %>%
  filter(parameter == "temperature") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(t_bmp180 = mean(value_hrf))

# bmp180 pressure hPa
press_bmp180 <- sensor_data %>%
  filter(sensor == "bmp180") %>%
  filter(parameter == "pressure") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(p_bmp180 = mean(value_hrf))

# Join Sensor Data by Nodes
j1 <- merge(temp_tsys01, temp_htu21d, by = c("node_id"), all = TRUE)
j2 <- merge(j1, temp_bmp180, by = c("node_id"), all = TRUE)
j3 <- merge(j2, press_bmp180, by = c("node_id"), all = TRUE)
j4 <- merge(j3, humid_htu21d, by = c("node_id"), all = TRUE)
node_data <- merge(j4, nodes, by = c("node_id"))

# Convert node data to spatial object
coordinates(node_data) <- node_data[,c("lon", "lat")]

# set data to the same projection
# proj4string(node.temps) <- CRS("+init=epsg:4326")
# Error in CRS("+init=epsg:4326") : NA
# https://gis.stackexchange.com/questions/387072/r-spcrs-returns-na
proj4string(node_data) <- CRS(sf::st_crs(4326)[[2]])

writeOGR(node_data, ".", "pres_all_sensors", driver = "ESRI Shapefile")
```


## Explore the Data in GeoDa

* Overview of the [GeoDa Cheat Sheet](http://geodacenter.github.io/cheatsheet.html)


* Generate box plots for each sensor

> Goal explore the distribution of each sensor.


* Generate scatter plots for each sensor

> Goal explore correlations between sensor data.

> Maybe the sensors are integrated and high pressure causes faulty reading on **tsys01**


It appears that node **001e0610e809** is not working correctly, and node **001e0611536c**
has no tsys01 sensor.


## Generate Shapefile for tsys01

**Choose the Hour for Node Data**

```{r final_node_data}
# Load Packages
library(tidyverse)
library(lubridate)
library(sp)
library(rgdal)

# Clear Workspace
rm(list = ls())

# Read Data
sensor_data <- read.csv("Data/data.csv.gz")
nodes <- read.csv("Data/nodes.csv")

# Fix timestamp so GeoDa doesn't crash on import
sensor_data$timestamp <- ymd_hms(sensor_data$timestamp)
nodes$start_timestamp <- ymd_hms(nodes$start_timestamp)
nodes$end_timestamp <- ymd_hms(nodes$end_timestamp)

# Filter Sensors by Sensor for hour
fhour <- 14 # filter hour

# tsys01 temperature C
temp_tsys01 <- sensor_data %>%
  filter(sensor == "tsys01") %>%
  filter(parameter == "temperature") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(t_tsys01 = mean(value_hrf))

# Remove outlier nodes 001e0610e809, and 001e0611536c
temp_tsys01 <- subset(temp_tsys01, node_id != "001e0610e809")
temp_tsys01 <- subset(temp_tsys01, node_id != "001e0611536c")

# Join Sensor Data by Nodes
node_data <- merge(temp_tsys01, nodes, by = c("node_id"))

# Convert node data to spatial object
coordinates(node_data) <- node_data[,c("lon", "lat")]

# set data to the same projection
# proj4string(node.temps) <- CRS("+init=epsg:4326")
# Error in CRS("+init=epsg:4326") : NA
# https://gis.stackexchange.com/questions/387072/r-spcrs-returns-na
proj4string(node_data) <- CRS(sf::st_crs(4326)[[2]])

writeOGR(node_data, ".", "pres_tsys01", driver = "ESRI Shapefile")
```


## Explore Final Temperature Data GeoDa


### Investigate Outliers and Patterns

* Generate box plot for tsys01

> Goal explore the distribution of tsys01.


* Generate a bubble map for tsys01

> Goal explore x y temp pattern


### Investigate Spatial Autocorrelation

* Space - Univariate Moran's I

> Goal determine spatial autocorrelation of temp nodes

**Generate Weights file GeoDa Tools > Weights Manager**

[L4: Spatial Autocorrelation](https://www.e-education.psu.edu/geog586/node/524)

*Weights file we need a representation of the contiguity structure of the map,*
*that is, which map units are neighbors to each other.*

*Cases in the upper-right quadrant indicate nodes temps, and local average node temps*
*are higher than overall average node temps.*

*Cases in the lower-left quadrand indicate node temps, and local average node temps*
*are lower than overall average node temps.*

**We expect temperature readings for sensors close to each other to be similar, and**
**temperature readings for sensors far from each other to be different.**

Because there is spatial autocorrelation with the tsys01 data we can use Kriging
to interpolate the temperature surface.


## Interpolate a Temperature Surface

**Choose the Hour for Node Data**

```{r krig, echo=FALSE}
library(sp) #spatial data wrangling & analysis
library(rgdal) #spatial data wrangling
library(rgeos) #spatial data wrangling & analytics
library(tidyverse) # data wrangling
library(tmap) #modern data visualizations
library(leaflet) #modern data visualizations
library(lubridate) #data wrangling
library(raster) #spatial raster data wrangling
library(gstat) #kriging and geostatistics

# Clear Workspace
rm(list = ls())

# Read Data
sensor_data <- read.csv("Data/data.csv.gz")
nodes <- read.csv("Data/nodes.csv")
chiCA <- readOGR("Data","ChiComArea", verbose = FALSE)

# Fix timestamp so GeoDa doesn't crash on import
sensor_data$timestamp <- ymd_hms(sensor_data$timestamp)
nodes$start_timestamp <- ymd_hms(nodes$start_timestamp)
nodes$end_timestamp <- ymd_hms(nodes$end_timestamp)

# Filter Sensors by Sensor for hour
fhour <- 14 # filter hour
ptitle <- "Exp Krig Avg Temp 2018-08-25 between 14:00 and 15:00"


# tsys01 temperature C
temp_tsys01 <- sensor_data %>%
  filter(sensor == "tsys01") %>%
  filter(parameter == "temperature") %>%
  filter(hour(timestamp) >= fhour) %>%
  filter(fhour >= hour(timestamp)) %>%
  group_by(node_id) %>%
  summarize(t_tsys01 = mean(value_hrf))

# Remove outlier nodes 001e0610e809, and 001e0611536c
temp_tsys01 <- subset(temp_tsys01, node_id != "001e0610e809")
temp_tsys01 <- subset(temp_tsys01, node_id != "001e0611536c")

# Join Sensor Data by Nodes
node_data <- merge(temp_tsys01, nodes, by = c("node_id"))

# Convert node data to spatial object
coordinates(node_data) <- node_data[,c("lon", "lat")]

# generate a grid from Chicago area
pt2grid <- function(ptframe,n) {
  bb <- bbox(ptframe)  
  ptcrs <- proj4string(ptframe)  
  xrange <- abs(bb[1,1] - bb[1,2])  
  yrange <- abs(bb[2,1] - bb[2,2])  
  cs <- c(xrange/n,yrange/n)  
  cc <- bb[,1] + (cs/2)  
  dc <- c(n,n)  
  x1 <- GridTopology(cellcentre.offset=cc,cellsize=cs,cells.dim=dc)  
  x2 <- SpatialGrid(grid=x1,proj4string=CRS(ptcrs))
  return(x2)
}

# plot 100 x 100 cell grid
chi.grid <- pt2grid((chiCA),100)

# set data to the same projection
# proj4string(node.temps) <- CRS("+init=epsg:4326")
# Error in CRS("+init=epsg:4326") : NA
# https://gis.stackexchange.com/questions/387072/r-spcrs-returns-na
proj4string(node_data) <- CRS(sf::st_crs(4326)[[2]])
proj4string(chiCA) <- CRS(sf::st_crs(4326)[[2]])
proj4string(chi.grid) <- CRS(sf::st_crs(4326)[[2]])

# variogram
tmp.vgm <- variogram(node_data$t_tsys01 ~ 1, node_data)

# plot exponential model
tmp.fit.exp<- fit.variogram(tmp.vgm, model=vgm("Exp"))

# prepare exponential
temp.kriged.exp <- krige(node_data$t_tsys01 ~ 1, node_data, chi.grid, model = tmp.fit.exp, debug.level = 0)
chi.temp.kriged.exp <- temp.kriged.exp[chiCA,]

# plot exponential krig
m.exp <- tm_shape(chi.temp.kriged.exp) +
  tm_raster("var1.pred", style = "jenks", title = "Temperature (C)", palette = "BuPu") + tm_shape(node_data) + tm_dots(size=0.01) +
  tm_layout(main.title = ptitle, main.title.size = 1.1) +
  tm_legend(position = c("left", "bottom"))

tmap_arrange(m.exp)
```
